---
title: "Exercise 4: Transportation Problem"
output: html_notebook
---

For this exercise, you will work on a problem that requires you to develop a delivery schedule. For this. you will use the transportation problem, presnted in the readings.

* Readings:
+ `05-Transportation-Problem.Rmd` (found under Labs/Assignments) 
+ `Reading: Transportation Problem.pdf`

* Due date:	See course outline for due date.

* Materials: This R Notebook, and the following RData file (found under Labs/Assignments)
+ A file called `hamilton_graph_zoned.RData`

* Instructions:	Answers must be provided by typing your solutions using RStudio and this R Notebook file. Change the name of the file to `Exercise-5 (Name-Student Number).Rmd`

Add new code chunkS by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*. When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

***

## Preliminaries

Begin by clearing the workspace and loading the datasets that you will need for this exercise. They are available from the Readings section. For instance, you will need the `rgdal` package to read the shape file. Make sure all packages that you need are installed and loaded `library()`. 
```{r echo=FALSE}
rm(list = ls())
```

Next, load the packages the you will use in this exercise.
```{r message=FALSE}
library(igraph)
library(scales)
library(sf)
library(tidygraph)
library(tidyverse)
library(units)
```

Load road network:
```{r}
load("hamilton_graph_zoned.RData")
```

The task at hand is as follows:

Paez Pop is a small craft root beer manufacturer. To preserve quality, the company produces their fair trade, organic, probiotic root beer in only two plants, A and B, where the secret ingredient is readily available. From these two plants, the root beer is shipped to four stores (Store I, II, III, and IV). Because of the increased cost of diesel, shipping cost has become a major expense for the company, so they have hired you to determine whether shipping costs can be reduced. 

The plant production managers have estimated the output from their respective plants, as follows:

* Plant A: output is 20 truckloads of root beer
* Plant B: output is 40 truckloads of root beer

The CEO of Paez Pop has allocated the total output to the given warehouses by examining sales data from prior months, as follows:

* Store I: 15 truckloads of root beer
* Store II: 10 truckloads of root beer
* Store III: 20 truckloads of root beer
* Store IV: 15 truckloads of root beer

The cost of transportation per truckload is $14/km:

The current delivery schedule that the company uses is shown in the table below. The unit of measurement is the truckload:

Plant/Distribution  | Store I | Store II  | Store III  | Store IV  | Supply
------------------- | ------- | ------    | --------   | --------  | ------
Plant A             | 10      | 5         | 0          |  5        | 20
Plant B             |  5      | 5         | 20         | 10        | 40
Demand              | 15      | 10        | 20         | 15

Your job is to determine the quantity (i.e., the number of truckloads) of root beer that should be shipped from each plant to each warehouse so that the company's total shipping cost is minimized.

***

# EXERCISE 5: Spatial Interaction Analysis (100 marks)

To begin, choose _one_ node from each column for the location of the two plants and the four stores:

Plant A | Plant B | Store I | Store II | Store III | Store IV
--------|---------|---------|----------|-----------|----------
  8072  |  17530  |   1097  |   14347  |   15975   |   14720  
 23557  |  18015  |  21230  |    9827  |   12802   |   12483
 23437  |  23749  |  15775  |   13954  |           |   21659
 17413  |         |         |          |           |

1. Find the shortest-path distance from the plants to the stores and obtain a cost matrix (use a freight rate of $14/km) **(10)**

>> Select nodes for plants and stores:

```{r}
plant_ids <- c(23557, 17530)
store_ids <- c(21230, 13954, 15975, 21659)
```

>> Plot:

```{r}
ggplot() +
  geom_sf(data = hamilton_graph_zoned %>%
            activate(edges) %>%
            as_tibble() %>%
            st_as_sf(),
          color = "lightgray") +
  geom_sf(data = hamilton_graph_zoned %>%
            activate(nodes) %>%
            as_tibble() %>%
            st_as_sf() %>%
            filter(nodeID %in% plant_ids),
          color = "red") +
  geom_sf(data = hamilton_graph_zoned %>%
            activate(nodes) %>%
            as_tibble() %>%
            st_as_sf() %>%
            filter(nodeID %in% store_ids),
          color = "blue")
```

>> To calculate the shortest paths the students need to add the length of the links to the road network:

```{r}
hamilton_graph_zoned <- hamilton_graph_zoned %>% # Pass the object `hamilton_graph_zoned` to the following function
  activate(edges) %>% # Activate the edges of the `tbl_graph` object
  mutate(length = st_length(geometry) %>% # Use function `st_length()` and the geometry of the edges to calculate the length of the link, and use `mutate()` to create a new column in the table to store this information
           set_units("km"))
```

>> Once link lengths are available, `igraph::distances()` is used to find the travel times on the shortest paths between zone centroids:

```{r}
node_distances <- distances(hamilton_graph_zoned, # Calculate shortest path costs in a graph
                            v = hamilton_graph_zoned %>% # Choose the origin nodes in the graph for calculating distances
                              activate(nodes) %>% # Activate the nodes
                              filter(nodeID %in% plant_ids) %>% # Filter nodes that are _not_ NAs. This will select only centroid nodes, i.e., those with a valid GTA06 identifier 
                              pull(nodeID), # Pull the node identifiers for the origins
                            to = hamilton_graph_zoned %>% # Choose the origin nodes in the graph for calculating distances
                              activate(nodes) %>% # Activate the nodes
                              filter(nodeID %in% store_ids) %>% # Filter nodes that are _not_ NAs. This will select only centroid nodes, i.e., those with a valid GTA06 identifier 
                              pull(nodeID), # Pull the node identifiers for the destinations
                            weights = hamilton_graph_zoned %>% # The weights will be the link lengths (i.e., distance)
                              activate(edges) %>% # Activate the edges of the graph
                              pull(length)) # Pull the length of the links
node_distances
```

>> The cost matrix is the distance times the per km rate

```{r}
cost_matrix <- node_distances * 14
cost_matrix
```

2. Calculate the cost of transportation according to the current delivery schedule **(10)**

>> Current delivery schedule

Plant/Distribution  | Store I | Store II  | Store III  | Store IV  | Supply
------------------- | ------- | ------    | --------   | --------  | ------
Plant A             | 10      | 5         | 0          |  5        | 20
Plant B             |  5      | 5         | 20         | 10        | 40
Demand              | 15      | 10        | 20         | 15

>> Write the current delivery schedule in matrix form:

```{r}
current_schedule <- c(10, 5, 0, 5, 5, 5, 20, 10) %>%
  matrix(nrow = 2, byrow = TRUE)
current_schedule
```

>> The total cost is sum of the products of deliveries by cost by route:

```{r}
sum(current_schedule * cost_matrix)
```

3. Create the initial simplex tableau for this problem **(10)**

>> This is the initial simplex tableau, copying the supply/demand constraints and the costs

```{r}
s1 <- c(-1, -1, -1, -1, 0, 0, 0, 0, 20)
s2 <- c(0, 0, 0, 0, -1, -1, -1, -1, 40)
s3 <- c(1, 0, 0, 0, 1, 0, 0, 0, -15)
s4 <- c(0, 1, 0, 0, 0, 1, 0, 0, -10)
s5 <- c(0, 0, 1, 0, 0, 0, 1, 0, -20)
s6 <- c(0, 0, 0, 1, 0, 0, 0, 1, -15)
x11 <- c(1, 0, 0, 0, 0, 0, 0, 0, 0)
x12 <- c(0, 1, 0, 0, 0, 0, 0, 0, 0)
x13 <- c(0, 0, 1, 0, 0, 0, 0, 0, 0)
x14 <- c(0, 0, 0, 1, 0, 0, 0, 0, 0)
x21 <- c(0, 0, 0, 0, 1, 0, 0, 0, 0)
x22 <- c(0, 0, 0, 0, 0, 1, 0, 0, 0)
x23 <- c(0, 0, 0, 0, 0, 0, 1, 0, 0)
x24 <- c(0, 0, 0, 0, 0, 0, 0, 1, 0)
P <- c(0, 0, 0, 0, 0, 0, 0, 0, 1)
Constant <- c(cost_matrix[1,], cost_matrix[2,], 0)
tableau <- cbind(s1, s2, s3, s4, s5, s6, x11, x12, x13, x14, x21, x22, x23, x24, P, Constant)
tableau
```

4. Solve the problem using the simplex method, and create a table showing the optimized flows that you have generated for their distribution system. What is the minimized cost, and how much savings have you found for the company? **(30)**

>> Create variables for the number of rows and number of columns:

```{r}
nr <- nrow(tableau) # number of rows in the tableau
nr

nc <- ncol(tableau) # number of columns in the tableau
nc
```

>> Copy the code from the reading with the solution to the transportation problem:

```{r}
stopCriterion <- min(tableau[nr,])
while(stopCriterion < 0){ # While the stopping criterion is still negative, do the following instructions
  pivotCol <- which(tableau[nr,] == min(tableau[nr,])) # Identify candidates for pivot column
  pivotCol <- pivotCol[1] # Break possible ties in candidates for pivot column
  candidateRows <- which(tableau[,pivotCol] == 1) # Find candidate rows for pivot row
  pivotRow <- candidateRows[which(tableau[candidateRows, nc] == min(tableau[candidateRows, nc]))] # Identify the pivot row
  nonZero <- which(tableau[, pivotCol] != 0) # Find non-zero elements in the pivot column
  nonZero <- nonZero[which(nonZero != pivotRow)] # Remove the pivot row from the list of rows with non-zero elements in the pivot column
  for(i in 1:length(nonZero)){ # For all rows that need to be updated, do the following instruction
    tableau[nonZero[i],] <- tableau[nonZero[i],] - tableau[nonZero[i], pivotCol] * tableau[pivotRow,] # Update the row to create unit column
  }
  stopCriterion <- min(tableau[nr,]) # Update the stopping criterion
}
tableau
```

>> Current delivery schedule

Plant/Distribution  | Store I | Store II  | Store III  | Store IV  | Supply
------------------- | ------- | ------    | --------   | --------  | ------
Plant A             | 5       | 0         | 0          | 15        | 20
Plant B             | 10      | 10        | 20         | 0         | 40
Demand              | 15      | 10        | 20         | 15

>> New schedule

Plant/Distribution  | Store I | Store II  | Store III  | Store IV  | Supply
------------------- | ------- | ------    | --------   | --------  | ------
Plant A             | 0       | 5         | 0          | 15        | 20
Plant B             | 15      | 5         | 20         | 0         | 40
Demand              | 15      | 10        | 20         | 15

>> Total cost: $19,424.70

>> Savings are the original cost minus the new cost:

```{r}
sum(current_schedule * cost_matrix) - tableau[nr, nc]
```

5. The key to consulting is getting repeat business from your clients, so you're happy when the CEO of Paez Pop calls you back months later to ask you to do some more work.

His company's root beer has become so successful that they now estimate potential demand at Store I and Store IV has increased by 10 truckloads of root beer a month, at each of the two stores. This extra demand has been going unmet, for now, because they haven't yet had the ability to expand production capacity at either plant, and haven't modified their distribution system since you optimized it for them in question 2.

Paez Pop has now saved up enough money to expand one, and only one, of their two plants by 20 truckloads of extra root beer production, so they can meet this new demand. The CEO knows the two plants have different transportation costs to the four stores, so he is wondering if you can tell him which of the two facilities he should expand, in order to minimize his company's transportation costs.

Find which plant, when its production is increased by 20, will find Paez Pop the higher savings. **(25)**

>> Case 1, plant A is expanded:

```{r}
u1 <- c(-1, -1, -1, -1, 0, 0, 0, 0, 40)
u2 <- c(0, 0, 0, 0, -1, -1, -1, -1, 40)
u3 <- c(1, 0, 0, 0, 1, 0, 0, 0, -25)
u4 <- c(0, 1, 0, 0, 0, 1, 0, 0, -10)
u5 <- c(0, 0, 1, 0, 0, 0, 1, 0, -20)
u6 <- c(0, 0, 0, 1, 0, 0, 0, 1, -25)
x11 <- c(1, 0, 0, 0, 0, 0, 0, 0, 0)
x12 <- c(0, 1, 0, 0, 0, 0, 0, 0, 0)
x13 <- c(0, 0, 1, 0, 0, 0, 0, 0, 0)
x14 <- c(0, 0, 0, 1, 0, 0, 0, 0, 0)
x21 <- c(0, 0, 0, 0, 1, 0, 0, 0, 0)
x22 <- c(0, 0, 0, 0, 0, 1, 0, 0, 0)
x23 <- c(0, 0, 0, 0, 0, 0, 1, 0, 0)
x24 <- c(0, 0, 0, 0, 0, 0, 0, 1, 0)
P <- c(0, 0, 0, 0, 0, 0, 0, 0, 1)
Constant <- c(cost_matrix[1,], cost_matrix[2,], 0)
tableau <- cbind(u1, u2, u3, u4, u5, u6, x11, x12, x13, x14, x21, x22, x23, x24, P, Constant)
tableau
```

>> the code again:

```{r}
stopCriterion <- min(tableau[nr,])
while(stopCriterion < 0){ # While the stopping criterion is still negative, do the following instructions
  pivotCol <- which(tableau[nr,] == min(tableau[nr,])) # Identify candidates for pivot column
  pivotCol <- pivotCol[1] # Break possible ties in candidates for pivot column
  candidateRows <- which(tableau[,pivotCol] == 1) # Find candidate rows for pivot row
  pivotRow <- candidateRows[which(tableau[candidateRows, nc] == min(tableau[candidateRows, nc]))] # Identify the pivot row
  nonZero <- which(tableau[, pivotCol] != 0) # Find non-zero elements in the pivot column
  nonZero <- nonZero[which(nonZero != pivotRow)] # Remove the pivot row from the list of rows with non-zero elements in the pivot column
  for(i in 1:length(nonZero)){ # For all rows that need to be updated, do the following instruction
    tableau[nonZero[i],] <- tableau[nonZero[i],] - tableau[nonZero[i], pivotCol] * tableau[pivotRow,] # Update the row to create unit column
  }
  stopCriterion <- min(tableau[nr,]) # Update the stopping criterion
}
tableau
```

>> Case 2, plant B is expanded:

```{r}
u1 <- c(-1, -1, -1, -1, 0, 0, 0, 0, 20)
u2 <- c(0, 0, 0, 0, -1, -1, -1, -1, 60)
u3 <- c(1, 0, 0, 0, 1, 0, 0, 0, -25)
u4 <- c(0, 1, 0, 0, 0, 1, 0, 0, -10)
u5 <- c(0, 0, 1, 0, 0, 0, 1, 0, -20)
u6 <- c(0, 0, 0, 1, 0, 0, 0, 1, -25)
x11 <- c(1, 0, 0, 0, 0, 0, 0, 0, 0)
x12 <- c(0, 1, 0, 0, 0, 0, 0, 0, 0)
x13 <- c(0, 0, 1, 0, 0, 0, 0, 0, 0)
x14 <- c(0, 0, 0, 1, 0, 0, 0, 0, 0)
x21 <- c(0, 0, 0, 0, 1, 0, 0, 0, 0)
x22 <- c(0, 0, 0, 0, 0, 1, 0, 0, 0)
x23 <- c(0, 0, 0, 0, 0, 0, 1, 0, 0)
x24 <- c(0, 0, 0, 0, 0, 0, 0, 1, 0)
P <- c(0, 0, 0, 0, 0, 0, 0, 0, 1)
Constant <- c(cost_matrix[1,], cost_matrix[2,], 0)
tableau <- cbind(u1, u2, u3, u4, u5, u6, x11, x12, x13, x14, x21, x22, x23, x24, P, Constant)
tableau
```

>> the code:

```{r}
stopCriterion <- min(tableau[nr,])
while(stopCriterion < 0){ # While the stopping criterion is still negative, do the following instructions
  pivotCol <- which(tableau[nr,] == min(tableau[nr,])) # Identify candidates for pivot column
  pivotCol <- pivotCol[1] # Break possible ties in candidates for pivot column
  candidateRows <- which(tableau[,pivotCol] == 1) # Find candidate rows for pivot row
  pivotRow <- candidateRows[which(tableau[candidateRows, nc] == min(tableau[candidateRows, nc]))] # Identify the pivot row
  nonZero <- which(tableau[, pivotCol] != 0) # Find non-zero elements in the pivot column
  nonZero <- nonZero[which(nonZero != pivotRow)] # Remove the pivot row from the list of rows with non-zero elements in the pivot column
  for(i in 1:length(nonZero)){ # For all rows that need to be updated, do the following instruction
    tableau[nonZero[i],] <- tableau[nonZero[i],] - tableau[nonZero[i], pivotCol] * tableau[pivotRow,] # Update the row to create unit column
  }
  stopCriterion <- min(tableau[nr,]) # Update the stopping criterion
}
tableau
```

>> Check which of the two plant expansions leads to a least cost

>> MARKS:

>> /10, 5 marks for constructing each new simplex tableau
>> /5 for knowing how to ctrl-c ctrl-v code twice
>> /5 for solving 19,700 and 20,900 for case 1 and case 2
>> /5 for determining that production facility 1 is the one that should be expanded.

>> Note: if either simplex tableau has an error, all subsequent answers may be wrong. So in essence, they're getting 20 marks for constructing 2 simplex tableaux and 5 marks for comparing two numbers.

6. Some time later, after expanding capacity at one of the plants, Paez Pop is thinking of trying to reduce costs by investing in a new, futuristic transportation trend: they're going to fire all their truck drivers at Plant B, and replace them with shiny new electric-powered autonomous vehicles from Space Karen They think this will save them 20% on all transport costs from plant B to each of the six stores.

(Plant A's truck drivers can't be replaced because they're unionized, so Plant A's transport costs will remain the same.)

By how many percent would this plan reduce Paez Pop's total transport costs in a new optimized network? What do the flows look like now? **(15)**

>> You only need to change the cost line of the tableau, which essentially means multiplying all costs from Plant B by 0.8 (this is the 20% reduction):

```{r}
u1 <- c(-1, -1, -1, -1, 0, 0, 0, 0, 40)
u2 <- c(0, 0, 0, 0, -1, -1, -1, -1, 40)
u3 <- c(1, 0, 0, 0, 1, 0, 0, 0, -25)
u4 <- c(0, 1, 0, 0, 0, 1, 0, 0, -10)
u5 <- c(0, 0, 1, 0, 0, 0, 1, 0, -20)
u6 <- c(0, 0, 0, 1, 0, 0, 0, 1, -25)
x11 <- c(1, 0, 0, 0, 0, 0, 0, 0, 0)
x12 <- c(0, 1, 0, 0, 0, 0, 0, 0, 0)
x13 <- c(0, 0, 1, 0, 0, 0, 0, 0, 0)
x14 <- c(0, 0, 0, 1, 0, 0, 0, 0, 0)
x21 <- c(0, 0, 0, 0, 1, 0, 0, 0, 0)
x22 <- c(0, 0, 0, 0, 0, 1, 0, 0, 0)
x23 <- c(0, 0, 0, 0, 0, 0, 1, 0, 0)
x24 <- c(0, 0, 0, 0, 0, 0, 0, 1, 0)
P <- c(0, 0, 0, 0, 0, 0, 0, 0, 1)
Constant <- c(cost_matrix[1,], 0.8 * cost_matrix[2,], 0)
tableau <- cbind(u1, u2, u3, u4, u5, u6, x11, x12, x13, x14, x21, x22, x23, x24, P, Constant)
tableau
```

>> ctrl-c ctrl-v:

```{r}
stopCriterion <- min(tableau[nr,])
while(stopCriterion < 0){ # While the stopping criterion is still negative, do the following instructions
  pivotCol <- which(tableau[nr,] == min(tableau[nr,])) # Identify candidates for pivot column
  pivotCol <- pivotCol[1] # Break possible ties in candidates for pivot column
  candidateRows <- which(tableau[,pivotCol] == 1) # Find candidate rows for pivot row
  pivotRow <- candidateRows[which(tableau[candidateRows, nc] == min(tableau[candidateRows, nc]))] # Identify the pivot row
  nonZero <- which(tableau[, pivotCol] != 0) # Find non-zero elements in the pivot column
  nonZero <- nonZero[which(nonZero != pivotRow)] # Remove the pivot row from the list of rows with non-zero elements in the pivot column
  for(i in 1:length(nonZero)){ # For all rows that need to be updated, do the following instruction
    tableau[nonZero[i],] <- tableau[nonZero[i],] - tableau[nonZero[i], pivotCol] * tableau[pivotRow,] # Update the row to create unit column
  }
  stopCriterion <- min(tableau[nr,]) # Update the stopping criterion
}
tableau
```

```{r}
1- (21527.840583/24417.006362)
```

>> Total transportation costs are reduced by 11.8%. Flows do not change at all.

>> /5 for the new simplex tableau - full marks if they only change the cost line then rebind
>> /5 ctrl-c ctrl-v and solve for 48040
>> /5 for calculate percent savings


